First Order Alternation

### 2018-10-19 quantifiers/draft.tex
ok

@@

Typically, when checking the unreachability of a set of program
configurations \cite{mcmillan06}, the interpolants used to annotate
the unfolded control structure are assertions about the values of the
program variables in a given control state, at a certain step of a
given execution. However, in an alternating model of computation, it
is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

@@

The main application of first-order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, the method described in
\S\ref{sec:emptiness} can extend our previous \emph{generic register
  automata} \cite{HolikIosifRogalewiczVojnar18} inclusion checking
framework, by allowing monitor (right-hand side) automata to have
local variables, that are not visible in the language\footnote{We
  conjecture that hidden computations using invisible variables
  (clocks) are the reason for which timed and register automata cannot
  be complemented within their class.}.

@@

Formally, a run of a predicate automaton $P =
\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\mu$, for each rule
$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
$\mu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
for all $q \in Q \setminus F$.

@@

\begin{proposition}
  Given a predicate automaton $P =
  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
  first-order alternating automaton that has the same description as
  $P$. Then $\lang{P} = \lang{\A_P}$. 
\end{proposition}
\proof{
\qed}
As before, this result enables using the semi-algorithm from
\S\ref{sec:emptiness} for checking emptiness of predicate automata.
We point out that, although quantifier-free predicate automata with
predicates of arity one are decidable for emptiness \cite{Farzan15},
currently there is no method for checking emptiness of predicate
automata with predicates of arity greater than one, other than the
explicit enumeration of cubes. Moreover, no method for dealing with
emptiness in the presence of transition quantifiers is known to exist.

>>>>

\section{Introduction}

Many results in formal language theory rely on the assumption that
languages are defined over finite alphabets. In practice, this
assumption is not realistic when attempting to model real-time
hardware and communication protocols or even simple programs, whose
input and observable output requires taking into account data values,
ranging over very large domains, better viewed as infinite
mathematical abstractions.

Alternating automata are a generalization of nondeterministic automata
with universal transitions, that create several copies of the
automaton, which synchronize on the same input word. Alternating
automata are appealing for verification because they allow encoding of
problems such as temporal logic model checking in linear time, as
opposed to the exponential time required by nondeterministic automata
\cite{Vardi95}.

%% However, checking the emptiness of an alternating automaton over a
%% finite alphabet is \pspace-complete, instead of
%% \nlogspace-complete, which is the case for nondeterministic automata.
%% In practice, this complexity can be dealt with using efficient
%% antichain algorithms \cite{Ganty09}.

A finite-alphabet alternating automaton is typically described by
transitions $q \arrow{a}{} \phi$, where $q$ is a state, $a$ is an
input symbol and $\phi$ is a positive boolean combinations of states,
viewed as propositional variables. Here we introduce a generalized
alternating automata model in which states are predicate symbols
$q(y_1,\ldots,y_k)$, the input has associated data variables
$x_1,\ldots,x_n$, ranging over an infinite domain and transitions are
of the form $q(y_1,\ldots,y_k) \arrow{a(x_1,\ldots,x_n)}{} \phi$,
where $\phi$ is any formula in the first-order theory of the data
domain, in which each state predicate occurs under an even number of
negations. In this model, the arguments of a predicate atom
$q(y_1,\ldots,y_k)$ track the values of the \emph{internal variables}
associated with the state. Together with the input values
$x_1,\ldots,x_n$, these values are used to compute the successor
states and are invisible in the input sequence.

Previous attempts to generalized classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory (register) automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This excludes the possibility of encoding a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$. However, when all internal
variables are made visible in the input, determinization and
complementation are possible, for data theories closed under negation,
such as integer or real arithmetics \cite{IosifRV16}.

The solution we adopt is a tight coupling of internal variables to
control states. As we show, this allows for linear-time
complementation just as in the case of boolean alternating
automata. Complementation is, moreover, possible when the transition
formulae contain first-order quantifiers, generating
infinitely-branching executions.

The price to be paid for this expressivity is that emptiness of
first-order alternating automata is undecidable, even for the simplest
data theory of equality \cite{Farzan15}. We work around this problem
by developing an abstraction refinement lazy annotation semi-algorithm
in the spirit of the \impact~procedure for checking safety of
nondeterministic integer programs \cite{McMillan06}. However, checking
emptiness of first-order alternating automata by lazy annotation with
interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules are problematic to decide if a given
  trace is spurious, mainly because adding uninterpreted predicate
  symbols to decidable first-order theories, such as Presburger
  arithmetic, results in undecidability \cite{Halpern91}. We deal with
  this problem by assuming that the first-order data theory, without
  uninterpreted predicate symbols, has an effective quantifier
  elimination procedure, that instantiates quantifers with witness
  terms.
%
\item The interpolants used to prove the spuriousness of an input
  sequence are not \emph{local} anymore, as they may refer to input
  values encountered in the past. However, the future executions are
  oblivious to when these values have been seen in the past and depend
  only on the data constraints between the values. We use this
  observation to define a labeling of nodes visited by the lazy
  annotation semi-algorithm as conjunctions of existentially
  quantified interpolants that combine predicate symbols with data
  constraints.
\end{compactenum}

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata. Further, we
describe applications of this method to several undecidable but
recursively enumerable problems, for which no effective
semi-algorithms exist, to date: inclusion between recognizable timed
languages \cite{AlurDill94} and between languages recognized by
finite-memory automata \cite{KaminskiFrancez94}, as well as emptiness
of predicate automata, a model of first-order alternating automata
used to check safety and liveness properties of parameterized
concurrent programs \cite{Farzan15,Farzan16}.

\paragraph{\bf Related Work} 
The first-order alternating automata model presented in this paper
stems from our previous work on boolean alternating automata extended
with variables ranging over infinite data \cite{IosifXu18}. There we
considered that states are propositional variables, as in the
classical textbook alternating automata model, and that all the
variables of the automaton are reflected in the input. The model in
this paper overcomes this latter restriction by allowing for internal
variables, whose variables are not visible in the language. This
solves an older language inclusion problem $\bigcap_{i=1}^n\lang{A_i}
\subseteq \lang{B}$, between finite-state automata with data variables
\cite{IosifRV16}. There, we assumed that all variables of the observer
automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
model the concurrent components of the system under check.

@@

Typically, when checking the unreachability of a set of program
configurations \cite{McMillan06}, the interpolants used to annotate
the unfolded control structure are assertions about the values of the
program variables in a given control state, at a certain step of a
given execution. However, in an alternating model of computation, it
is useful to distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are defined by
formulae involving a single predicate symbol, whereas
\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
and variables $\stamp{x}{i}$, for a single $i \geq 0$.

@@

The main application of first-order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, the method described in
\S\ref{sec:emptiness} can extend our previous \emph{generic register
  automata} \cite{IosifRV16} inclusion checking framework, by allowing
monitor (right-hand side) automata to have local variables, that are
not visible in the language.

@@

Formally, a run of a predicate automaton $P =
\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\nu$, for each rule
$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
$\nu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
for all $q \in Q \setminus F$.

@@

\begin{proposition}
  Given a predicate automaton $P =
  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
  first-order alternating automaton that has the same description as
  $P$. Then $\lang{P} = \lang{\A_P}$. 
\end{proposition}
\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
  \lang{P}$ be a word and $\I_0, \ldots, \I_n$ be an accepting
  execution of $P$ over $w$. Let $\stamp{\I}{i}_j$ be the
  interpretation that associates each predicate $\stamp{q}{i}$ the set
  $\I_j(q)$, for $i,j \in [n]$. Then one builds, by induction on
  $n\geq0$, an execution $\T$ of $\A_P$ such that $\I_\T \subseteq
  \bigcup_{i=0}^n \stamp{\I}{i}_i$, where $\I_\T$ is the unique
  interpretation associated with $\T$. Since $\I_0, \ldots, \I_n$ is
  accepting, we have $\stamp{\I}{n}_n(\stamp{q}{n}) = \emptyset$, for
  all $q \in Q \setminus F$ and hence $\I_\T(\stamp{q}{n}) =
  \emptyset$, for all $q \in Q \setminus F$ and, consequently $w \in
  \lang{\A_P}$. ``$\supseteq$'' Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n) \in \lang{\A_P}$ be a word and $\T$ be an accepting
  execution of $\A_P$ over $w$. We define the sequence of
  interpretations $\I_0, \ldots, \I_n$ as $\I_i(q) =
  \I_\T(\stamp{q}{i})$, for each $i \in [n]$ and each $q \in Q$. By
  induction on $n\geq0$ one shows that $\I_0, \ldots, \I_n$ is an
  execution $P$. Moreover, since $\T$ is accepting, we have $\I_n(q) =
  \I_\T(\stamp{q}{n}) = \emptyset$, for each $q \in Q \setminus F$,
  thus $w \in \lang{P}$. \qed}

As before, this result enables using the
semi-algorithm from \S\ref{sec:emptiness} for checking emptiness of
predicate automata.  We point out that, although quantifier-free
predicate automata with predicates of arity one are decidable for
emptiness \cite{Farzan15}, currently there is no method for checking
emptiness of predicate automata with predicates of arity greater than
one, other than the explicit enumeration of cubes. Moreover, no method
for dealing with emptiness in the presence of transition quantifiers
is known to exist.
